---
title: "20220913_FPvsPD_hte_TMT16_peptide"
author: "Tianen He"
date: "2022/9/13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(xlsx, magrittr)
```

# Input files

## PD without contaminants

```{r}
pd_protein = read.csv("./TMT16_PD3/TMT16plex_chimerys_intensity_nocontam_220914_Proteins.txt", sep = "\t")
pd_peptide = read.csv("./TMT16_PD3/TMT16plex_chimerys_intensity_nocontam_220914_PeptideGroups.txt", sep = "\t")
```

## FP with contaminants

```{r}
fp_protein_tmt = read.csv("./TMT16_FP18/tmt-report/abundance_protein_None.tsv", sep = "\t")
fp_peptide_tmt = read.csv("./TMT16_FP18//tmt-report/abundance_peptide_None.tsv", sep = "\t")
```

## contaminant

```{r}
contaminant_map = read.csv("./uniprot_contaminant_maps.txt", sep = "\t")
contaminant_entry = contaminant_map$Entry

channel_num = 4*15
```

# PD

## protein

```{r}
colnames(pd_protein)
pd_protein_abundance = pd_protein[,c(96:110,112:126,128:142,144:158)-65] # 4*15 samples, without pool
colnames(pd_protein_abundance)
```

### Proteins with an abundance of NA in all 15*4 channels are considered as unquantified in PD

```{r}
# proteins with 15*4 NA are removed
pd_protein_abundance_quantified = pd_protein_abundance %>% 
    magrittr::extract(rowSums(is.na(.)) != channel_num, )
```

### calculate missing rate

```{r}
# remove the completely not quantified proteins
pd_protein_num_quantified = dim(pd_protein_abundance_quantified)[1]
pd_na_count_quantified = length(pd_protein_abundance_quantified[pd_protein_abundance_quantified==0]) # include NAs and zeros
pd_missing_rate_quantified = pd_na_count_quantified/(pd_protein_num_quantified * channel_num)
pd_missing_rate_quantified
```

## peptide

```{r}
# colnames(pd_peptide)
pd_peptide_abundance = pd_peptide[,c(96:110,112:126,128:142,144:158)-79] # 4*15 samples, without pool
colnames(pd_peptide_abundance)

# proteins with 15*4 NA are removed
pd_peptide_abundance_quantified = pd_peptide_abundance %>% 
    extract(rowSums(is.na(.)) != channel_num, )
```

# FP

## protein

### tmt-report (abundance) (remove contaminants) (4 batches)

```{r}
# remove contaminants
fp_protein_tmt_real = fp_protein_tmt[!(fp_protein_tmt$Index %in% contaminant_entry), ]
fp_protein_num_tmt_real = dim(fp_protein_tmt_real)[1]

# extract abundance    
fp_protein_abundance_tmt_real = fp_protein_tmt_real[, 6:65] # 4 batches
colnames(fp_protein_abundance_tmt_real)
```

### calculate quantified (all are quantified)

```{r}
# proteins with 15*4 NA are removed
tmp = fp_protein_abundance_tmt_real
tmp[is.na(tmp)] = 0
summary((apply(tmp, 1, sum))!=0)
```

### calculate missing rate

```{r}
# tmt-report  (after removal of contaminants)
fp_na_count_tmt_real = length(fp_protein_abundance_tmt_real[fp_protein_abundance_tmt_real==0]) # include NAs and zeros
fp_missing_rate_tmt_real = fp_na_count_tmt_real/(fp_protein_num_tmt_real * channel_num)
fp_missing_rate_tmt_real
```

## peptide

### tmt-report (abundance) (remove contaminants)

```{r}
fp_peptide_tmt_real = fp_peptide_tmt[!(fp_peptide_tmt$ProteinID %in% contaminant_entry), ]
fp_peptide_num_tmt_real = dim(fp_peptide_tmt_real)[1]
```
