---
title: "20220913_FPvsPD_hte_TMT16"
author: "Tianen He"
date: "2022/9/13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(xlsx, VennDiagram, pheatmap, ggplot2, tidyr, ggsignif, limma,
               clusterProfiler, org.Hs.eg.db, magrittr, stringr,
               ggsci, ggthemes, ggVennDiagram, sva)

colors = pal_jco()(10)
hcolors = pal_simpsons()(16)

pd1 = "#1e88e5"
pd2 = "#bbdefb"

fp1 = "#ffb300"
fp2 = "#ffecb3"

root = "D:/hetianen/Documents/Guomics/FPvsPD paper 2.0/PD3.0/"

channel_num = 15 * 4

sample_infos = read.xlsx("../cvdnx_kidney_label20210801_20211124edited.xlsx", sheetIndex = 1)
```

# input files

variable pd, fp: quantified proteins without contaminants
fp: input files are in log2 scale

```{r}
pd_unquant = read.csv("./TMT16_PD3/TMT16plex_chimerys_intensity_nocontam_220914_Proteins.txt", sep = "\t") 

fp_contam = read.csv("./TMT16_FP18/tmt-report/ratio_protein_None.tsv", sep = "\t") # contain contaminants
fp_abundance_contam = read.csv("./TMT16_FP18/tmt-report/abundance_protein_None.tsv", sep = "\t") # contain contaminants 

contaminant_map = read.csv(paste0(root, "uniprot_contaminant_maps.txt"), sep = "\t")
contaminant_entry = contaminant_map$Entry

# batch-free data for limma are imported below
```

# process protein matrix

```{r}
# remove unquantified proteins
tmp = pd_unquant[, c(96:110,112:126,128:142,144:158)-65]
pd = pd_unquant[which(rowSums(is.na(tmp))!=dim(tmp)[2]), ]
rm(tmp, pd_unquant)

# remove contaminant
fp = fp_contam[which(!(fp_contam$Index %in% contaminant_entry)),]
fp_abundance_all = fp_abundance_contam[which(!(fp_abundance_contam$Index %in% contaminant_entry)),]
rm(fp_contam, fp_abundance_contam)

# sort by accession
pd_sorted = pd[order(pd$Accession), ]
fp_sorted = fp[order(fp$Index), ]
fp_abundance_sorted = fp_abundance_all[order(fp_abundance_all$Index), ]

# extract log2 abundances and linear ratios (60 channels)
pd_abundance = pd_sorted[, c(96:110,112:126,128:142,144:158)-65] # abundance
pd_abundance = log2(pd_abundance) # original output data are linear-scaled
rownames(pd_abundance) = pd_sorted$Accession

# calculate abundance ratio
tmp_sample = pd_sorted[, c(96:110,112:126,128:142,144:158)-65]
tmp_reference = pd_sorted[, rep(c(95,111,127,143), each = 15)-65]
# tmp_sample[is.na(tmp_sample)] = 0
tmp_reference[is.na(tmp_reference)] = 0
tmp_ratio = tmp_sample/tmp_reference %>% data.frame()
tmp_ratio[sapply(tmp_ratio,is.infinite)] = 100
pd_ratio = tmp_ratio
rm(tmp_ratio, tmp_sample, tmp_reference)
rownames(pd_ratio) = pd_sorted$Accession

# colnames(fp_sorted)
fp_ratio = fp_sorted[, 6:65] # original output data are log2-scaled
fp_ratio = 2^fp_ratio
rownames(fp_ratio) = fp_sorted$Index

# colnames(fp_abundance_sorted)
fp_abundance = fp_abundance_sorted[, 6:65] # original output data are log2-scaled
rownames(fp_abundance) = fp_sorted$Index

rm(pd_sorted, fp_sorted)
```

## unify column names and orders

```{r}
# FP
## rename columns
samples = unlist(strsplit(colnames(fp_ratio), "X"))[seq(2, 120, 2)]
samples = data.frame(V1 = unlist(strsplit(samples, "_[2-5]$")), 
                     V2 = rep(c(2, 3, 4, 5), each = 15))
samples = paste0("b", as.numeric(samples$V2) - 1, "_", samples$V1) # b2-b5 -> b1-b4
colnames(fp_ratio) = samples
colnames(fp_abundance) = colnames(fp_ratio)
colnames(pd_ratio) = colnames(fp_ratio)
colnames(pd_abundance) = colnames(fp_ratio)

```

```{r}
# write.csv(pd_ratio, file = paste0(root, "PD_ratio_tmt16.csv"), quote = FALSE)
# write.csv(pd_abundance, file = paste0(root, "PD_abundance[log2]_tmt16.csv"), quote = FALSE)

# write.csv(fp_ratio, file = paste0(root, "FP_ratio_tmt16.csv"), quote = FALSE)
# write.csv(fp_abundance, file = paste0(root, "FP_abundance[log2]_tmt16.csv"), quote = FALSE)
```

# sample grouping infos

### technical replicates

(old batch number)
b5_133C b3_133N control cortex 
b3_133C b2_129C covid medulla
b2_133C b4_128N covid cortex

```{r}
reps = list()
reps$group1 = c("b2_133N", "b4_133C")
reps$group2 = c("b1_129C", "b2_133C")
reps$group3 = c("b3_128N", "b1_133C")
```

```{r}
# 53 channels (available in sample info table)
sample_id = sample_infos$MS_Sample.ID %>% 
    str_replace(., "b2", "b1") %>% 
    str_replace(., "b3", "b2") %>% 
    str_replace(., "b4", "b3") %>% 
    str_replace(., "b5", "b4")
sample_group = sample_infos$Group # COVID, Control
sample_tissue = sample_infos$Tissue %>% 
    gsub(pattern = " medulla", replacement = "Medulla") %>% 
    gsub(pattern = " cortex", replacement = "Cortex") # Cortex, Medulla
sample_infos$Tissue = sample_tissue
sample_batch = substring(sample_id, first = 1, last = 2)

# 60 channels
groups = rep("Other", 60)
groups[match(sample_id, samples)] = sample_group
tissues = rep("Other", 60)
tissues[match(sample_id, samples)] = sample_tissue
# batches = rep(c("b2", "b3", "b4", "b5"), each = 15)
batches = rep(c("b1", "b2", "b3", "b4"), each = 15)

## replicates
for (i in names(reps)) {
    groups[match(reps[[i]][2], samples)] = groups[match(reps[[i]][1], samples)]
    tissues[match(reps[[i]][2], samples)] = tissues[match(reps[[i]][1], samples)]
}
```

## extract 53 labeled channels

```{r}
pd_abundance_labeled = pd_abundance[, match(sample_id, samples)]
pd_ratio_labeled = pd_ratio[, match(sample_id, samples)]
fp_abundance_labeled = fp_abundance[, match(sample_id, samples)]
fp_ratio_labeled = fp_ratio[, match(sample_id, samples)]
```

# overlap of proteins quantified by two softwares

```{r}
pd_accession = pd$Accession
fp_accession = fp$Index
both_accession = sort(intersect(pd_accession, fp_accession))
```

```{r}
venn = venn.diagram(list("PD" = pd_accession, "FP" = fp_accession), 
             # paste0(root, "Figure2A_protein_overlap.svg"),
             filename = NULL,
             # imagetype = "svg",
             height = 800, width = 800,
             alpha = 0.6, lwd = 1, cex = 1,
             cat.cex = 0.8, margin = 0.1, cat.dist = 0.07,
             ext.text = TRUE,
             ext.percent = c(0.2, 0.2, 0.2),
             # fill = c("dodgerblue", "goldenrod1")
             fill = c(pd1, fp1),
             ext.length = 0.85
             )

ggsave(venn, path = root,
       file = "Figure2A_protein_overlap.pdf", device = "pdf",
       width = 600, height = 600, units = "px")
```

## comparison of abundance between overlapped and non-overlapped proteins (PD and FP merged)

PD and FP merged to one plot

```{r}
# data1, data2: protein matrix, include only abundances/ratios
# both_accession: accessions of overlapped proteins
# ylabel: eg. abundance, intensity, log2 abundance
compare_overlap_nonoverlap_merged = function (data1, data2, both_accession, 
                                              device = "pdf", ylabel) {
    
    # preprocess data
    num_channels = dim(data1)[2]
    data1$software = rep("PD", dim(data1)[1])
    data1$overlap = rep("unique", dim(data1)[1])
    data1[which(rownames(data1) %in% both_accession), ]$overlap = "overlapping"
    data1$overlap = as.factor(data1$overlap)
    
    data2$software = rep("FP", dim(data2)[1])
    data2$overlap = rep("unique", dim(data2)[1])
    data2[which(rownames(data2) %in% both_accession), ]$overlap = "overlapping"
    data2$overlap = as.factor(data2$overlap)
    
    data = rbind(data1, data2)
    data_longer = pivot_longer(data, cols = c(1:num_channels), names_to = "channel")
    data_longer$facet = str_c(data_longer$software, data_longer$overlap, sep = " ")
    data_longer$batch = data_longer$channel %>% 
        str_extract(., pattern = "^b.+_") %>% 
        # Perl regex: "." means any chars except \n, "+" means repeat 1 or more times
        str_sub(., end = -2) %>% # remove "_"
        as.factor()
    num_batch = data_longer$batch %>% 
        levels() %>% length()
    
    p_template_violin = ggplot(data_longer, aes(x=facet, y=value)) +
        geom_violin(aes(fill=facet)) +
        geom_signif(comparisons = list(c("PD overlapping", "PD unique"),
                                       c("FP overlapping", "FP unique")),
                    # y_position = 32,
                    test = wilcox.test, textsize = 3, map_signif_level = TRUE) +
        ylab(ylabel) + xlab("") +
        theme_few() +
        # scale_fill_simpsons() +
        scale_fill_manual(values=c(fp1, fp2, pd1, pd2)) +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              legend.title = element_blank())
    
    # one violin plot for each batch
    p_compare_batch_violin = p_template_violin +
        facet_wrap(~batch, ncol = num_batch)
    
    # one violin plot for each channel
    p_compare_channel_violin = p_template_violin +
        facet_wrap(~channel, nrow = 5)
    
    return(list(p_compare_batch_violin, p_compare_channel_violin))
}
```

```{r}
violins = compare_overlap_nonoverlap_merged(
    pd_abundance, fp_abundance, both_accession, ylabel = "Log2 (Protein abundance)")

p_compare_batch_violin = violins[[1]]
p_compare_channel_violin = violins[[2]]

ggsave(file = paste0(root, "Figure2C_protein_overlap_compare_batch_violin.pdf"), 
       plot = p_compare_batch_violin, width = 2400, height = 1000, units = "px")

ggsave(file = paste0(root, "FigureS1_protein_overlap_compare_channel_violin.pdf"),
       plot = p_compare_channel_violin, width = 3000, height = 3000, units = "px")      
```

# correlation coefficients

## correlation of abundance ratios of each sample (53 channels)

```{r}
# normalize
# pd_ratio_scaled = t(scale(t(pd_ratio_labeled)))
# fp_ratio_scaled = t(scale(t(fp_ratio_labeled)))
pd_ratio_scaled = pd_ratio_labeled
fp_ratio_scaled = fp_ratio_labeled

# proteins identified in common
pd_ratio_both = pd_ratio_scaled[both_accession, ] %>% as.data.frame()
fp_ratio_both = fp_ratio_scaled[both_accession, ] %>% as.data.frame()

colnames(pd_ratio_both) = paste("PD", colnames(pd_ratio_both), sep = "_")
colnames(fp_ratio_both) = paste("FP", colnames(fp_ratio_both), sep = "_")

cor_matrix = cor(pd_ratio_both, fp_ratio_both, use = "pairwise.complete.obs",
                 method = "spearman")
# cor_pd = cor(pd_ratio_both, use = "pairwise.complete.obs") # for row clustering
# cor_fp = cor(fp_ratio_both, use = "pairwise.complete.obs") # for column clustering
```

```{r}
# anno_row = data.frame(Group = sample_group, Tissue = sample_tissue,
#                       Batch = sample_batch, row.names = colnames(pd_ratio_both))
# anno_col = data.frame(Group = sample_group, Tissue = sample_tissue,
#                       Batch = sample_batch, row.names = colnames(fp_ratio_both))
# 
# anno_color = list(Batch = c(b1 = hcolors[1], b2 = hcolors[5], 
#                             b3 = hcolors[4], b4 = hcolors[14]),
#                   Tissue = c(Cortex = hcolors[9], Medulla = hcolors[6]),
#                   Group = c(Control = hcolors[7], COVID = hcolors[10]))
# 
# pheatmap(cor_matrix, 
#          cluster_cols = F, cluster_rows = F,
#          # clustering_distance_rows = as.dist(1 - cor_pd),
#          # clustering_distance_cols = as.dist(1 - cor_fp),
#          filename = paste0(root, "protein_cor_matrix_53samples_tmt16.pdf"), 
#          annotation_col = anno_col,
#          annotation_row = anno_row,
#          annotation_colors = anno_color,
#          # main = "Pearson Correlation Matrix",
#          color = cividis(100, direction = 1),
#          width = 10, height = 9, fontsize = 8,
#          # show_colnames = F, show_rownames = F,
#          # fontsize_row = 6, fontsize_col = 6,
#          # angle_col = "315"
#          )
```

```{r}
# extract diagonal
cor_self = diag(cor_matrix)
cor_self_median = median(cor_self)
cor_self_median

hist_plot = ggplot(data.frame(cor_self), aes(x = cor_self, y = ..count..)) +
    geom_histogram(bins = 10, fill = colors[10], color = "white") +
    # ggtitle("Distribution of Pearson Correlation Coefficient") + 
    xlab("Spearman's correlation coefficient") +
    ylab("Count of samples") +
    # ylim(0, 20) +
    theme_few() +
    geom_vline(xintercept = cor_self_median, color = colors[2], size = 1, linetype = 1)
ggsave(plot = hist_plot, file = paste0(
    root, "FigureS2F_correlation_coefficient_distribution_53samples_tmt16.pdf"),
       width = 1200, height = 1000, units = "px")
hist_plot
```

## correlation of abundance ratios of each protein (53 channels)

```{r}
# transpose
pd_ratio_t = t(pd_ratio_labeled) %>% as.data.frame() 
fp_ratio_t = t(fp_ratio_labeled) %>% as.data.frame() 

# extract proteins in common
pd_ratio_both_t = pd_ratio_t[, sort(both_accession)] %>% as.data.frame()
fp_ratio_both_t = fp_ratio_t[, sort(both_accession)] %>% as.data.frame()

cor_self_t = list()
for (prot in both_accession) {
    cor_t = cor(pd_ratio_both_t[[prot]], fp_ratio_both_t[[prot]], 
                method = "spearman", use = "pairwise.complete.obs")
    cor_self_t[[prot]] = cor_t
}
cor_self_t = unlist(cor_self_t)

summary(cor_self_t)
cor_self_median_t = median(cor_self_t, na.rm = T)
cor_self_median_t

hist_plot_t = ggplot(data.frame(cor_self_t), aes(x = cor_self_t, y = ..count..)) +
    geom_histogram(bins = 20, fill = colors[10], color = "white") +
    # ggtitle("Distribution of Pearson Correlation Coefficient") + 
    xlab("Spearman's correlation coefficient") +
    ylab("Count of proteins") +
    # ylim(0, 20) +
    theme_few() +
    geom_vline(xintercept = cor_self_median_t, color = colors[2], size = 1, linetype = 1)
ggsave(plot = hist_plot_t, file = paste0(root, "Figure2D_correlation_coefficient_distribution_53samples_proteins_tmt16.pdf"),
       width = 1200, height = 1000, units = "px")
hist_plot_t
```

## correlation of protein matrix as a whole

```{r}
pd_ratio_both_vector = unlist(pd_ratio_both)
fp_ratio_both_vector = unlist(fp_ratio_both)

cor = cor(pd_ratio_both_vector, fp_ratio_both_vector, use = "pairwise.complete.obs",
          method = "spearman")
cor.test(pd_ratio_both_vector, fp_ratio_both_vector, use = "pairwise.complete.obs",
         alternative = "two.sided", method = "spearman")
```

## correlation of technical replicates

### PD

```{r}
pd_abundance_both_60 = pd_abundance

pd_cor_reps = list()
pd_cor_reps_test = list()
for (i in names(reps)) {
    quant_rep1 = pd_abundance_both_60[reps[[i]][1]] %>% unlist()
    quant_rep2 = pd_abundance_both_60[reps[[i]][2]] %>% unlist()
    pd_cor_reps[[i]] = cor(quant_rep1, quant_rep2, use = "pairwise.complete.obs",
                           method = "spearman")
    pd_cor_reps_test[[i]] = cor.test(quant_rep1, quant_rep2, 
                                     alternative = "two.sided", 
                                     method = "spearman")
}
pd_cor_reps
pd_cor_reps_test
```

### FP

```{r}
fp_abundance_both_60 = fp_abundance

fp_cor_reps = list()
fp_cor_reps_test = list()
for (i in names(reps)) {
    quant_rep1 = fp_abundance_both_60[reps[[i]][1]] %>% unlist()
    quant_rep2 = fp_abundance_both_60[reps[[i]][2]] %>% unlist()
    fp_cor_reps[[i]] = cor(quant_rep1, quant_rep2, use = "pairwise.complete.obs",
                           method = "spearman")
    fp_cor_reps_test[[i]] = cor.test(quant_rep1, quant_rep2,
                                     alternative = "two.sided",
                                     method = "spearman")
}
fp_cor_reps
fp_cor_reps_test
```

### t-test 

```{r}
t.test(unlist(fp_cor_reps), unlist(pd_cor_reps), alternative = "greater")
```

# ---------------------------------------------------

# preprocess protein matrix for limma

```{r}
median_normalize = function (data) {
    # check median centering
    par(mfrow = c(1, 2))
    boxplot(data, las = 2, main = "before")
    
    # median normalization by channel (make the median of each channel zero)
    data_medians = apply(data, 2, median, na.rm = TRUE)
    data_norm = as.data.frame(scale(data, center = data_medians, scale = FALSE))
    
    # check median centering
    boxplot(data_norm, las = 2, main = "after")
    
    return(data_norm)
}
```

## imputation, log2 transformation, normalization

```{r}
pd_limma = pd_ratio_labeled
fp_limma = fp_ratio_labeled

# 220919 added: replace zeros to avoid errors downstream
pd_limma[pd_limma == 0] = NA
fp_limma[fp_limma == 0] = NA

# impute NAs with 0.8 of minimum
pd_limma[is.na(pd_limma)] = 0.8*min(pd_limma, na.rm = T)
fp_limma[is.na(fp_limma)] = 0.8*min(fp_limma, na.rm = T)

# log2 transformation
pd_limma = log2(pd_limma)
fp_limma = log2(fp_limma)

# median normalization
pd_limma_norm = median_normalize(pd_limma)
fp_limma_norm = median_normalize(fp_limma)
```

```{r}
# save data for BatchServer
# write.csv(pd_limma_norm, file = "./PD_ratio_53_limma_batch.csv", quote = FALSE)
# write.csv(fp_limma_norm, file = "./FP_ratio_53_limma_batch.csv", quote = FALSE)
```

## (BatchServer) evaluate, remove, and evaluate again batch effect

### alternative: ComBat

```{r}
# pd_limma_norm = read.csv(paste0(root, "PD_ratio_53_limma_batch.csv"), row.names = 1)
# fp_limma_norm = read.csv(paste0(root, "FP_ratio_53_limma_batch.csv"), row.names = 1)

mod = model.matrix(~as.factor(Group) +as.factor(Tissue), data = sample_infos)

# pd_batchfree = ComBat(dat = pd_limma_norm, batch = sample_batch, par.prior = F, mod = mod)
# fp_batchfree = ComBat(dat = fp_limma_norm, batch = sample_batch, par.prior = F, mod = mod)

# write.csv(pd_batchfree, file = paste0(root, "PD_ratio_53_limma_batchfree_combat.csv"), quote = FALSE)
# write.csv(fp_batchfree, file = paste0(root, "FP_ratio_53_limma_batchfree_combat.csv"), quote = FALSE)

# pd_batchfree = read.csv(paste0(root, "PD_ratio_53_limma_batchfree_combat.csv"), row.names = 1)
# fp_batchfree = read.csv(paste0(root, "FP_ratio_53_limma_batchfree_combat.csv"), row.names = 1)
```

## median normalize batch-free data

```{r}
# import data
# pd_batchfree = read.csv(paste0(root, "PD_ratio_53_limma_batchfree.csv"), row.names = 1)
# fp_batchfree = read.csv(paste0(root, "FP_ratio_53_limma_batchfree.csv"), row.names = 1)
```

```{r}
# median normalization by channel
pd_bf_norm = median_normalize(pd_batchfree)
fp_bf_norm = median_normalize(fp_batchfree)
```

```{r}
# save data for BatchServer
# write.csv(pd_bf_norm, file = paste0(root, "PD_ratio_53_limma_batchfree_norm.csv"), quote = FALSE)
# write.csv(fp_bf_norm, file = paste0(root, "FP_ratio_53_limma_batchfree_norm.csv"), quote = FALSE)
```

## (BatchServer) evaluate batch effect in batch-free data

# DE analysis with limma

## make design and contrasts

```{r}
# make design table
limma_group = interaction(sample_group, sample_tissue)
design = model.matrix(~0 + limma_group)
colnames(design) = levels(limma_group)
rownames(design) = sample_id

# make contrasts
x = c("COVID.Cortex-Control.Cortex", "COVID.Medulla-Control.Medulla")
contrast = makeContrasts(contrasts=x, levels=design)
```

## DEP analysis and plotting

```{r}
# contrast_which: column number or column name specifying which contrast of the linear model is of interest
# software, group: for output file naming, eg. PD, FP; cortex, medulla
limma_DEP = function(data, design, contrast_all, contrast_which, software, group,
                     p_thres = 0.05, lfc_thres = log2(1.5), # default parameter
                     p_thres_label = 0.05, lfc_thres_label = log2(1.5),
                     device = "pdf") {
    # fit
    fit1 = lmFit(data, design)
    head(coef(fit1))
    fit2 = contrasts.fit(fit1, contrasts = contrast_all)
    fit3 = eBayes(fit2)
    
    # plotSA(fit3, main="Final model: Mean-variance trend")
    
    # filter (default: fc > 1.5, adj.p < 0.05)
    dep_table = topTable(fit3, sort.by = "P", number = Inf, 
                         coef = contrast_which, 
                         p.value = p_thres, lfc = lfc_thres)
    dep_table$protein = rownames(dep_table)
    dep_table = dep_table[ , c("protein", colnames(dep_table)[1:6])]
    
    # map the UNIPROT id to gene symbol (remove duplicate)
    tmp = bitr(dep_table$protein, OrgDb = org.Hs.eg.db, drop = FALSE,
               fromType = "UNIPROT", toType = "SYMBOL")
    dup = duplicated(tmp$UNIPROT)
    
    for (i in 1:length(dup)) {
        j = length(dup) - i + 1 # from tail to head
        if (dup[j] == TRUE) {
            cat_symbol = paste0(tmp[j - 1, "SYMBOL"], ",", tmp[j, "SYMBOL"])
            tmp[j - 1, "SYMBOL"] = cat_symbol
        }
    }
    dep_table$symbol = tmp[which(!duplicated(tmp$UNIPROT)), "SYMBOL"]
    rm(tmp, dup)
    
    # save table
    write.table(dep_table, row.names = FALSE, sep = "\t", quote = FALSE,
                file = paste0(root, "dep_", software, "_", group, ".tsv"))
    
    # volcano plot
    dep_table_volcano = topTable(fit3, sort.by = "P", number = Inf, coef = contrast_which)
    dep_table_volcano$protein = rownames(dep_table_volcano)
    dep_table_volcano$threshold = ifelse(dep_table_volcano$adj.P.Val < p_thres & 
                                             abs(dep_table_volcano$logFC) >= lfc_thres,
                                         ifelse(dep_table_volcano$logFC >= lfc_thres, 
                                                'up', 'down'), 
                                         '-') %>% as.factor()
    
    # map the UNIPROT id to gene symbol (remove duplicate)
    tmp = bitr(dep_table_volcano$protein, OrgDb = org.Hs.eg.db, drop = FALSE,
               fromType = "UNIPROT", toType = "SYMBOL")
    dup = duplicated(tmp$UNIPROT)
    
    for (i in 1:length(dup)) {
        j = length(dup) - i + 1 # from tail to head
        if (dup[j] == TRUE) {
            cat_symbol = paste0(tmp[j - 1, "SYMBOL"], ",", tmp[j, "SYMBOL"])
            tmp[j - 1, "SYMBOL"] = cat_symbol
        }
    }
    dep_table_volcano$symbol = tmp[which(!duplicated(tmp$UNIPROT)), "SYMBOL"]
    rm(tmp, dup)
    
    # label specific points with SYMBOL
    dep_table_volcano$label = ifelse(dep_table_volcano$adj.P.Val < p_thres_label & 
                                     abs(dep_table_volcano$logFC) >= lfc_thres_label,
                                         dep_table_volcano$symbol, "")
    
    ## save table
    write.table(dep_table_volcano, row.names = FALSE, sep = "\t", quote = FALSE,
                file = paste0(root, "volcano_", software, "_", group, ".tsv"))
    
}
```

```{r}
lfc_thres = log2(1.5)
# PD
# voom(pd_bf_norm, design, plot = T)
limma_DEP(pd_bf_norm, design, contrast, contrast_which = 1,
          lfc_thres = lfc_thres, p_thres_label = 0, 
          device = "pdf", 
          software = "PD", group = "cortex")
limma_DEP(pd_bf_norm, design, contrast, contrast_which = 2,
          lfc_thres = lfc_thres, p_thres_label = 0,
          device = "pdf",
          software = "PD", group = "medulla")

# FP
limma_DEP(fp_bf_norm, design, contrast, contrast_which = 1,
          lfc_thres = lfc_thres, p_thres_label = 0,
          device = "pdf",
          software = "FP", group = "cortex")
limma_DEP(fp_bf_norm, design, contrast, contrast_which = 2,
          lfc_thres = lfc_thres, p_thres_label = 0,
          device = "pdf",
          software = "FP", group = "medulla")
```

## overlap of DEPs

```{r}
# import dep tables
dep_tables = list()
dep_tables$pd_cortex = read.table("dep_PD_cortex.tsv", header = T, sep = "\t")
dep_tables$pd_medulla = read.table("dep_PD_medulla.tsv", header = T, sep = "\t")
dep_tables$fp_cortex = read.table("dep_FP_cortex.tsv", header = T, sep = "\t")
dep_tables$fp_medulla = read.table("dep_FP_medulla.tsv", header = T, sep = "\t")

# DEP entries
dep_ids = list()
dep_ids$cortex$pd_up = dep_tables$pd_cortex %>% filter(logFC > 0) %>% extract2("protein")
dep_ids$cortex$pd_down = dep_tables$pd_cortex %>% filter(logFC < 0) %>% extract2("protein")
dep_ids$cortex$fp_up = dep_tables$fp_cortex %>% filter(logFC > 0) %>% extract2("protein")
dep_ids$cortex$fp_down = dep_tables$fp_cortex %>% filter(logFC < 0) %>% extract2("protein")

dep_ids$medulla$pd_up = dep_tables$pd_medulla %>% filter(logFC > 0) %>% extract2("protein")
dep_ids$medulla$pd_down = dep_tables$pd_medulla %>% filter(logFC < 0) %>% extract2("protein")
dep_ids$medulla$fp_up = dep_tables$fp_medulla %>% filter(logFC > 0) %>% extract2("protein")
dep_ids$medulla$fp_down = dep_tables$fp_medulla %>% filter(logFC < 0) %>% extract2("protein")
```

### venn diagrams

```{r}
# cortex
dep_cortex_venn = venn.diagram(list("PD up" = dep_ids$cortex$pd_up,
                  "PD down" = dep_ids$cortex$pd_down,
                  "FP up" = dep_ids$cortex$fp_up,
                  "FP down" = dep_ids$cortex$fp_down),
             # paste0(root, "dep_overlap_cortex.tiff"),
             filename = NULL,
             alpha = 0.6,
             height = 1200, width = 1400, cex = 1, cat.cex = 0.8,
             margin = 0.1,
             # cat.pos = c(200, 160),
             fill = c(pd1, pd2, fp1, fp2),
             main = "DEP cortex")
ggsave(dep_cortex_venn,
       path = root, filename = "Figure4A_dep_overlap_cortex.pdf", device = "pdf",
       width = 700, height = 700, units = "px")

# medulla
dep_medulla_venn = venn.diagram(list("PD up" = dep_ids$medulla$pd_up,
                  "PD down" = dep_ids$medulla$pd_down,
                  "FP up" = dep_ids$medulla$fp_up,
                  "FP down" = dep_ids$medulla$fp_down),
             # paste0(root, "dep_overlap_medulla.tiff"),
             filename = NULL,
             alpha = 0.6,
             height = 1200, width = 1400, cex = 1, cat.cex = 0.8,
             # cat.pos = c(200, 160),
             # fill = c("dodgerblue", "goldenrod1"),
             margin = 0.1,
             # fill = hcolors[c(1, 2, 5, 10)],
             fill = c(pd1, pd2, fp1, fp2),
             main = "DEP medulla")
ggsave(dep_medulla_venn,
       path = root, filename = "Figure4A_dep_overlap_medulla.pdf", device = "pdf",
       width = 700, height = 700, units = "px")
```

## enrichment analysis of DEPs

```{r}
enrich_analysis = function(dep_table, p_thres = 0.05, software, group) {
    # make a sorted gene list with uniprot id and FC
    dep_table_fc = 2^dep_table[, "logFC"]
    names(dep_table_fc) = as.character(dep_table[["protein"]])
    dep_table_fc = sort(dep_table_fc, decreasing = TRUE)
    
    # GO MF
    gomf = enrichGO(dep_table[["protein"]], OrgDb = org.Hs.eg.db,
                  ont = "MF", keyType = "UNIPROT", pvalueCutoff = p_thres)
    # write.table(gomf, file = paste0("go_mf_dep_", software, "_", group, ".tsv"),
    #             quote = FALSE, sep = "\t", row.names = FALSE)
    
    # GO BP
    gobp = enrichGO(dep_table[["protein"]], OrgDb = org.Hs.eg.db,
                  ont = "BP", keyType = "UNIPROT", pvalueCutoff = p_thres)
    # write.table(gobp, file = paste0("go_bp_dep_", software, "_", group, ".tsv"),
    #             quote = FALSE, sep = "\t", row.names = FALSE)
    
    # GO CC
    gocc = enrichGO(dep_table[["protein"]], OrgDb = org.Hs.eg.db,
                  ont = "CC", keyType = "UNIPROT", pvalueCutoff = p_thres)
    # write.table(gocc, file = paste0("go_cc_dep_", software, "_", group, ".tsv"),
    #             quote = FALSE, sep = "\t", row.names = FALSE)
    
    # KEGG
    kegg = enrichKEGG(dep_table[["protein"]], organism = 'hsa',
                      keyType = "uniprot", pvalueCutoff = p_thres)
    # write.table(kegg, file = paste0("kegg_dep_", software, "_", group, ".tsv"),
    #             quote = FALSE, sep = "\t", row.names = FALSE)
    
    return(list(gomf, gobp, gocc, kegg))
}
```

```{r}
go_result_list = list()
kegg_result_list = list()

go_id_list = list()
kegg_id_list = list()

for (software in c("PD", "FP")) {
    for (group in c("cortex", "medulla")) {
        acronym = ifelse(software == "FP", "fp", "pd")
        name = paste0(acronym, "_", group)
        table = dep_tables[[name]]
        
        return = enrich_analysis(table, software = software, group = group)
        
        go_result_list[[name]] = list()
        go_result_list[[name]][["MF"]] = return[[1]]
        go_result_list[[name]][["BP"]] = return[[2]]
        go_result_list[[name]][["CC"]] = return[[3]]
        kegg_result_list[[name]] = return[[4]]
        
        go_id_list[[name]] = list()
        go_id_list[[name]][["MF"]] = return[[1]]$ID
        go_id_list[[name]][["BP"]] = return[[2]]$ID
        go_id_list[[name]][["CC"]] = return[[3]]$ID
        kegg_id_list[[name]] = return[[4]]$ID
    }
}
```

```{r}
# save(go_result_list, kegg_result_list, go_id_list, kegg_id_list,
#      file = paste0(root, "go_kegg_results.RData"))
load(paste0(root, "go_kegg_results.RData"))
```

### overlap of enriched terms (venn)

```{r}
# cortex
for (type in c("BP", "MF", "CC")) {
    
    go_cortex_venn = venn.diagram(list("PD" = go_id_list$pd_cortex[[type]],
                                   "FP" = go_id_list$fp_cortex[[type]]),
             # paste0(root, "GO_overlap_cortex.tiff"),
             filename = NULL,
             height = 800, width = 800,
             alpha = 0.6, cex = 1, cat.cex = 0.8,
             main.cex = 0.8,
             margin = 0.2, cat.dist = 0.07,
             # cat.pos = c(320, 40),
             fill = c(pd1, fp1),
             main = paste0("Enriched GO ", type, "s in the cortex"))
    ggsave(go_cortex_venn,
           path = root,
           filename = paste0("Figure4C-S3A_GO_", type, "_overlap_cortex.pdf"), 
           units = "px", width = 600, height = 600)
}

kegg_cortex_venn = venn.diagram(list("PD" = kegg_id_list$pd_cortex,
                  "FP" = kegg_id_list$fp_cortex),
             # paste0(root, "KEGG_overlap_cortex.tiff"),
             filename = NULL,
             height = 800, width = 800,
             alpha = 0.6, cex = 1, cat.cex = 0.8,
             main.cex = 0.8,
             margin = 0.2, cat.dist = 0.07,
             # cat.pos = c(320, 40),
             fill = c(pd1, fp1),
             main = "Enriched KEGG pathways in the cortex")
ggsave(kegg_cortex_venn,
       path = root,
       filename = "Figure4C_KEGG_overlap_cortex.pdf", units = "px",
       width = 600, height = 600)

# medulla
for (type in c("BP", "MF", "CC")) {
    
    go_medulla_venn = venn.diagram(list("PD" = go_id_list$pd_medulla[[type]],
                                   "FP" = go_id_list$fp_medulla[[type]]),
             # paste0(root, "GO_overlap_cortex.tiff"),
             filename = NULL,
             height = 800, width = 800,
             alpha = 0.6, cex = 1, cat.cex = 0.8,
             main.cex = 0.8,
             margin = 0.2, cat.dist = 0.07,
             # cat.pos = c(320, 40),
             fill = c(pd1, fp1),
             main = paste0("Enriched GO ", type, "s in the medulla"))
    ggsave(go_medulla_venn,
           path = root,
           filename = paste0("FigureS3C_GO_", type, "_overlap_medulla.pdf"), 
           units = "px", width = 600, height = 600)
}

kegg_medulla_venn = venn.diagram(list("PD" = kegg_id_list$pd_medulla,
                  "FP" = kegg_id_list$fp_medulla),
             # paste0(root, "KEGG_overlap_cortex.tiff"),
             filename = NULL,
             height = 800, width = 800,
             alpha = 0.6, cex = 1, cat.cex = 0.8,
             main.cex = 0.8,
             margin = 0.2, cat.dist = 0.07,
             # cat.pos = c(320, 40),
             fill = c(pd1, fp1),
             main = "Enriched KEGG pathways in the medulla")
ggsave(kegg_medulla_venn,
       path = root,
       filename = "FigureS3C_KEGG_overlap_medulla.pdf", units = "px",
       width = 600, height = 600)
```

### draw comparison plots

```{r}
p_thres = 0.05

# cortex
dep_ids_cortex = list()
dep_ids_cortex$PD = c(dep_ids$cortex$pd_up, dep_ids$cortex$pd_down)
dep_ids_cortex$FP = c(dep_ids$cortex$fp_up, dep_ids$cortex$fp_down)

## go bp
compare_gobp_cortex = compareCluster(geneClusters = dep_ids_cortex, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "BP", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gobp_cortex, showCategory = 10) +
    ggtitle("Top enriched biological processes in the cortex") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "Figure4D_compare_gobp_cortex.pdf", units = "px",
       width = 1600, height = 1400)

## go mf (supplementary)
compare_gomf_cortex = compareCluster(geneClusters = dep_ids_cortex, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "MF", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gomf_cortex, showCategory = 10) +
    ggtitle("Top enriched GO MFs in the cortex") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 45))
ggsave(filename = "FigureS3B_compare_gomf_cortex.pdf", units = "px",
       width = 1500, height = 1800)

## go cc (supplementary)
compare_gocc_cortex = compareCluster(geneClusters = dep_ids_cortex, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "CC", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gocc_cortex, showCategory = 10) +
    ggtitle("Top enriched GO CCs in the cortex") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "FigureS3B_compare_gocc_cortex.pdf", units = "px",
       width = 1500, height = 1800)

## kegg
compare_kegg_cortex = compareCluster(geneClusters = dep_ids_cortex, fun = enrichKEGG,
                                     organism = 'hsa',
                                     keyType = "uniprot", pvalueCutoff = p_thres)
dotplot(compare_kegg_cortex, showCategory = 10) +
    ggtitle("Top enriched KEGG pathways in the cortex") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "Figure4D_compare_kegg_cortex.pdf", units = "px",
       width = 1600, height = 1400)
```


```{r}
# medulla
dep_ids_medulla = list()
dep_ids_medulla$PD = c(dep_ids$medulla$pd_up, dep_ids$medulla$pd_down)
dep_ids_medulla$FP = c(dep_ids$medulla$fp_up, dep_ids$medulla$fp_down)


## go bp
compare_gobp_medulla = compareCluster(geneClusters = dep_ids_medulla, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "BP", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gobp_medulla, showCategory = 10) +
    ggtitle("Top enriched GO BPs in the medulla") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "FigureS3D_compare_gobp_medulla.pdf", units = "px",
       width = 1500, height = 1800)

## go mf
compare_gomf_medulla = compareCluster(geneClusters = dep_ids_medulla, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "MF", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gomf_medulla, showCategory = 10) +
    ggtitle("Top enriched GO MFs in the medulla") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "FigureS3D_compare_gomf_medulla.pdf", units = "px",
       width = 1500, height = 1800)

## go cc
compare_gocc_medulla = compareCluster(geneClusters = dep_ids_medulla, fun = enrichGO,
                                   OrgDb = org.Hs.eg.db,
                                   ont = "CC", keyType = "UNIPROT", 
                                   pvalueCutoff = p_thres)
dotplot(compare_gocc_medulla, showCategory = 10) +
    ggtitle("Top enriched GO CCs in the medulla") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "FigureS3D_compare_gocc_medulla.pdf", units = "px",
       width = 1500, height = 1800)

## kegg
compare_kegg_medulla = compareCluster(geneClusters = dep_ids_medulla, fun = enrichKEGG,
                                     organism = 'hsa',
                                     keyType = "uniprot", pvalueCutoff = p_thres)
dotplot(compare_kegg_medulla, showCategory = 3) +
    ggtitle("Top enriched KEGG pathways in the medulla") +
    theme_few() +
    scale_y_discrete(labels = function(x) str_wrap(x, width = 50))
ggsave(filename = "FigureS3D_compare_kegg_medulla.pdf", units = "px",
       width = 1500, height = 1800)
```

### check the number in parentheses on the plots

```{r}
select(org.Hs.eg.db, dep_ids_cortex$PD, "GO", "UNIPROT") %>% 
    as.data.frame %>% .[which(.$ONTOLOGY == "BP"), ] %>%
    extract2("UNIPROT") %>% unique %>% length

select(org.Hs.eg.db, dep_ids_cortex$FP, "GO", "UNIPROT") %>% 
    as.data.frame %>% .[which(.$ONTOLOGY == "BP"), ] %>%
    extract2("UNIPROT") %>% unique %>% length
```
