---
title: "20220913_FPvsPD_hte_TMT6"
author: "Tianen He"
date: "2022/9/13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(xlsx, VennDiagram, tidyr, stringr, magrittr, ggplot2, pheatmap,
               ggsignif, ggsci, ggthemes, viridis, ggvenn, ggVennDiagram)

colors = pal_jco()(10)
hcolors = pal_simpsons()(16)

pd1 = "#1e88e5"
pd2 = "#bbdefb"

fp1 = "#ffb300"
fp2 = "#ffecb3"

root = "D:/hetianen/Documents/Guomics/FPvsPD paper 2.0/PD3.0/"
```

# Input files

## PD without contaminants

```{r}
pd_protein = read.csv("./TMT6_PD3/CPS2_chimerys_intensity_nocontam_hte220912_Proteins.txt", sep = "\t")
pd_peptide = read.csv("./TMT6_PD3/CPS2_chimerys_intensity_nocontam_hte220912_PeptideGroups.txt", sep = "\t")
```

## FP with contaminants

```{r}
fp_protein_tmt = read.csv("./TMT6_FP18/tmt-report/abundance_protein_None.tsv", sep = "\t")
fp_peptide_tmt = read.csv("./TMT6_FP18/tmt-report/abundance_peptide_None.tsv", sep = "\t")
```

## contaminant

```{r}
contaminant_map = read.csv("./uniprot_contaminant_maps.txt", sep = "\t")
contaminant_entry = contaminant_map$Entry

channel_num = 6
```

# PD

## protein

```{r}
# colnames(pd_protein)
pd_protein_abundance = pd_protein[,30:35]
colnames(pd_protein_abundance)
```

### Proteins with an abundance of NA in all 6 channels are considered as unquantified in PD

```{r}
# proteins with 6 NA are removed
pd_protein_abundance_quantified = pd_protein_abundance %>% 
    magrittr::extract(rowSums(is.na(.)) != channel_num, )

pd_protein_abundance_quantified_ids = pd_protein %>% 
    magrittr::extract(rowSums(is.na(pd_protein_abundance)) != channel_num, "Accession")
rownames(pd_protein_abundance_quantified) = pd_protein_abundance_quantified_ids
```

### calculate missing rate

```{r}
# remove the completely not quantified proteins
pd_protein_num_quantified = dim(pd_protein_abundance_quantified)[1]
pd_na_count_quantified = length(pd_protein_abundance_quantified[pd_protein_abundance_quantified==0]) # include NAs and zeros
pd_missing_rate_quantified = pd_na_count_quantified/(pd_protein_num_quantified * channel_num)
pd_missing_rate_quantified

pd_protein_abundance_quantified = log2(pd_protein_abundance_quantified)
```

## peptide

```{r}
# colnames(pd_peptide)
pd_peptide_abundance = pd_peptide[,16:21]
colnames(pd_peptide_abundance)
```

### calculate quantified

```{r}
# proteins with 6 NAs are removed
pd_peptide_abundance_quantified = pd_peptide_abundance %>% 
    extract(rowSums(is.na(.)) != channel_num, )
```

# FP

## protein

### tmt-report (abundance) (remove contaminants) 

```{r}
# remove contaminants
fp_protein_tmt_real = fp_protein_tmt[!(fp_protein_tmt$Index %in% contaminant_entry), ]
fp_protein_num_tmt_real = dim(fp_protein_tmt_real)[1]

# extract abundance    
# colnames(fp_protein_tmt_real)
fp_protein_abundance_tmt_real = fp_protein_tmt_real[, 6:11] 
colnames(fp_protein_abundance_tmt_real)

rownames(fp_protein_abundance_tmt_real) = fp_protein_tmt_real$Index
```

### calculate missing rate

```{r}
# tmt-report  (after removal of contaminants)
fp_na_count_tmt_real = length(fp_protein_abundance_tmt_real[fp_protein_abundance_tmt_real==0]) # include NAs and zeros
fp_missing_rate_tmt_real = fp_na_count_tmt_real/(fp_protein_num_tmt_real * channel_num)
fp_missing_rate_tmt_real
```

## peptide

### tmt-report (abundance) (remove contaminants)

```{r}
fp_peptide_tmt_real = fp_peptide_tmt[!(fp_peptide_tmt$ProteinID %in% contaminant_entry), ]
fp_peptide_num_tmt_real = dim(fp_peptide_tmt_real)[1]
```

# Sort rows, rename columns, save log2 abundance to files

```{r}
colnames(fp_protein_abundance_tmt_real) = 126:131
colnames(pd_protein_abundance_quantified) = colnames(fp_protein_abundance_tmt_real)

# sort by accession
pd_sorted = pd_protein_abundance_quantified[order(rownames(pd_protein_abundance_quantified)), ]
fp_sorted = fp_protein_abundance_tmt_real[order(rownames(fp_protein_abundance_tmt_real)), ]

# write.csv(pd_sorted, file = paste0(root, "PD_abundance[log2]_tmt6.csv"), quote = FALSE)
# write.csv(fp_sorted, file = paste0(root, "FP_abundance[log2]_tmt6.csv"), quote = FALSE)
```

# Overlap of proteins quantified by PD and FP

```{r}
pd_accession = rownames(pd_sorted)
fp_accession = rownames(fp_sorted)

both_accession = intersect(pd_accession, fp_accession)

venn = venn.diagram(list("PD" = pd_accession, "FP" = fp_accession), 
             filename = NULL,
             imagetype = "svg",
             height = 800, width = 800, 
             alpha = 0.5, lwd = 1, cex = 1, 
             cat.cex = 0.8, margin = 0.1, cat.dist = 0.07,
             ext.text = TRUE, ext.percent = c(0.2, 0.2, 0.2),
             fill = c(pd1, fp1),
             # cat.pos = c(70, 290),
             # ext.length = 1
             )
ggsave(venn, path = root,
       filename = "FigureS2A_protein_overlap_tmt6.pdf", units = "px",
       width = 600, height = 600)

```

## comparison of abundance between overlapped and non-overlapped proteins (PD and FP merged)

PD and FP merged to one plot

```{r}
# data1, data2: protein matrix, include only abundances/ratios
# both_accession: accessions of overlapped proteins
# ylabel: eg. abundance, intensity, log2 abundance
compare_overlap_nonoverlap_merged = function (data1, data2, both_accession, 
                                              device = "pdf", ylabel) {
    
    # preprocess data
    num_channels = dim(data1)[2]
    data1$software = rep("PD", dim(data1)[1])
    data1$overlap = rep("unique", dim(data1)[1])
    data1[which(rownames(data1) %in% both_accession), ]$overlap = "overlapping"
    data1$overlap = as.factor(data1$overlap)
    
    data2$software = rep("FP", dim(data2)[1])
    data2$overlap = rep("unique", dim(data2)[1])
    data2[which(rownames(data2) %in% both_accession), ]$overlap = "overlapping"
    data2$overlap = as.factor(data2$overlap)
    
    data = rbind(data1, data2)
    data_longer = pivot_longer(data, cols = c(1:num_channels), names_to = "channel")
    data_longer$facet = str_c(data_longer$software, data_longer$overlap, sep = " ")

    # one violin plot for each batch
    p_template_violin = ggplot(data_longer, aes(x=facet, y=value)) +
        geom_violin(aes(fill=facet)) +
        geom_signif(comparisons = list(c("PD overlapping", "PD unique"),
                                       c("FP overlapping", "FP unique")),
                    # y_position = 32,
                    test = wilcox.test, textsize = 3, map_signif_level = TRUE) +
        ylab(ylabel) + xlab("") +
        theme_few() +
        # scale_fill_simpsons() +
        scale_fill_manual(values=c(fp1, fp2, pd1, pd2)) +
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              legend.title = element_blank())
    
    return(p_template_violin)
}
```

```{r}
p_compare_batch_violin = compare_overlap_nonoverlap_merged(
    pd_sorted, fp_sorted, 
    both_accession, ylabel = "Log2 (Protein abundance)")
ggsave(file = paste0(root, "FigureS2C_protein_overlap_compare_batch_violin_PDandFP_tmt6.pdf"), 
       plot = p_compare_batch_violin, width = 1600, height = 1000, units = "px")
```

# Correlation coefficients

## correlation of log2 abundances of each sample (6 channels)

```{r}
# pd_ratio_scaled = scale(pd_sorted, scale = FALSE) # log2 abundance
# fp_ratio_scaled = scale(fp_sorted, scale = FALSE)
# pd_ratio_scaled = t(scale(t(pd_sorted))) # log2 abundance
# fp_ratio_scaled = t(scale(t(fp_sorted)))
pd_ratio_scaled = pd_sorted
fp_ratio_scaled = fp_sorted

# proteins identified in common
pd_ratio_both = pd_ratio_scaled[sort(both_accession), ] %>% as.data.frame()
fp_ratio_both = fp_ratio_scaled[sort(both_accession), ] %>% as.data.frame()

colnames(pd_ratio_both) = paste("PD", colnames(pd_ratio_both), sep = "_")
colnames(fp_ratio_both) = paste("FP", colnames(fp_ratio_both), sep = "_")

cor_matrix = cor(pd_ratio_both, fp_ratio_both, use = "pairwise.complete.obs",
                 method = "spearman")

pheatmap(cor_matrix, 
         cluster_cols = F, cluster_rows = F,
         filename = paste0(root, "protein_cor_matrix_tmt6.pdf"), 
         # main = "Pearson Correlation Matrix",
         color = cividis(100, direction = 1),
         width = 4, height = 3.6, fontsize = 8,
         # show_colnames = F, show_rownames = F,
         # fontsize_row = 6, fontsize_col = 6,
         # angle_col = "315"
         )

# extract diagonal
cor_self = diag(cor_matrix)
cor_self_median = median(cor_self)
cor_self_median

hist_plot = ggplot(data.frame(cor_self), aes(x = cor_self, y = ..count..)) +
    geom_histogram(bins = 10, fill = colors[10], color = "white") +
    # ggtitle("Distribution of Pearson Correlation Coefficient") + 
    xlab("Spearman's correlation coefficient") +
    ylab("Count of samples") +
    # ylim(0, 20) +
    theme_few() +
    geom_vline(xintercept = cor_self_median, color = colors[2], size = 1, linetype = 1)
ggsave(plot = hist_plot, 
       file = paste0(root, "FigureS2E_correlation_coefficient_distribution_tmt6_samples.pdf"),
       width = 1200, height = 1000, units = "px")
```

## correlation of log2 abundances of each protein

```{r}
# transpose
pd_ratio_t = t(pd_sorted) %>% as.data.frame() 
fp_ratio_t = t(fp_sorted) %>% as.data.frame() 

# extract proteins in common
pd_ratio_both_t = pd_ratio_t[, sort(both_accession)] %>% as.data.frame()
fp_ratio_both_t = fp_ratio_t[, sort(both_accession)] %>% as.data.frame()

# colnames(pd_ratio_both_t) = paste("PD", colnames(pd_ratio_both_t), sep = "_")
# colnames(fp_ratio_both_t) = paste("FP", colnames(fp_ratio_both_t), sep = "_")

cor_self_t = list()
for (prot in both_accession) {
    cor_t = cor(pd_ratio_both_t[[prot]], fp_ratio_both_t[[prot]], method = "spearman", 
                use = "pairwise.complete.obs")
    cor_self_t[[prot]] = cor_t
}
cor_self_t = unlist(cor_self_t)

summary(cor_self_t)
cor_self_median_t = median(cor_self_t)
cor_self_median_t

hist_plot_t = ggplot(data.frame(cor_self_t), aes(x = cor_self_t, y = ..count..)) +
    geom_histogram(bins = 20, fill = colors[10], color = "white") +
    # ggtitle("Distribution of Pearson Correlation Coefficient") + 
    xlab("Spearman's correlation coefficient") +
    ylab("Count of proteins") +
    # ylim(0, 20) +
    theme_few() +
    geom_vline(xintercept = cor_self_median_t, color = colors[2], size = 1, linetype = 1)
ggsave(plot = hist_plot_t, file = paste0(
    root, "FigureS2D_correlation_coefficient_distribution_tmt6_proteins.pdf"),
       width = 1200, height = 1000, units = "px")
hist_plot_t
```

## correlation of protein matrix as a whole

```{r}
pd_ratio_both_vector = unlist(pd_ratio_both)
fp_ratio_both_vector = unlist(fp_ratio_both)

cor = cor(pd_ratio_both_vector, fp_ratio_both_vector, use = "pairwise.complete.obs",
          method = "spearman")
cor.test(pd_ratio_both_vector, fp_ratio_both_vector,
         alternative = "two.sided", method = "spearman")
```